### 2021-09-28 documents/sections/extensions.tex
started adding PTir forms
added imperative data, some optics

\ex<bak-33>\bakairi \parencite[][4]{meira2003bakairi}\\
\begingl
\gla w-i-də//
\glb \gl{1}\gl{s_a_}-bathe-\gl{imm}//
\glft \qu{I bathed}//
\endgl
\xe
%
While \textcite[4]{meira2003bakairi} does list \obj{ge} \qu{to say}, \obj{tə} \qu{to go}, and \obj{əe(wɨ)} \qu{to come} as examples of \gl{s_a_} verbs, he does not indicate whether they belong to the class of \gl{s_a_}-1 verbs, with first person \obj{k-}, or the \gl{s_a_}-2 verbs, with \obj{w-}.\footnote{It should be noted at this point that \textcite{meira2003bakairi} indicates that the same verbs which take first person \obj{w-} in \bakairi also take a \gl{1+2} marker \obj{k-}. However, this marker is only illustrated for \qu{to bathe}, both by \textcite{meira2003bakairi} and \textcite{von1892bakairi}. Given the lack of data for other verbs, we will not further discuss this potential additional pattern. If the characterization by \citeauthor{meira2003bakairi} is accurate, then the pattern is fully parallel to the distribution of the first person prefixes.}
Luckily, while \textcite{von1892bakairi} did not accurately record all phonemic distinctions in \bakairi \parencite{meira2005bakairi}, he does provide inflected forms of cognates to the \arara verbs in \exref{arairr}.
We present them according to the phonemic analysis of \bakairi by \textcites{wheatley1969bakairi}{meira2003bakairi}{meira2005bakairi} in \exref{bakverbs}.

\pex[everyglpreamble=]<bakverbs> \bakairi \parencite[][131, 397, 76, 137, 374, 130]{von1892bakairi}
\begin{multicols}{2}
\a<bak-3>
\begingl
\glpreamble \ort{u-ɣépa} //
\gla u-ge-pa//
\glb \gl{1}\gl{s_a_}-say-\gl{neg}//
\glft \qu{I don't say.}//
\endgl
\a<bak-5>
\begingl
\glpreamble \ort{wi-táki} / \ort{wi-tági} //
\gla w-i-taki//
\glb \gl{1}\gl{s_a_}-be-\gl{int}//
\glft \qu{I was.}//
\endgl
\a<bak-4>
\begingl
\glpreamble \ort{kχaewí-le} //
\gla k-əewɨ-lɨ//
\glb \gl{1}\gl{s_a_}-come-\gl{imm}//
\glft \qu{I came.}//
\endgl
\a<bak-6>
\begingl
\glpreamble \ort{kχ-itaké-he} //
\gla k-ɨtəgɨ-se//
\glb \gl{1}\gl{s_a_}-go.down-\gl{npst}?//
\glft \qu{I go down.}//
\endgl
\a<bak-2>
\begingl
\glpreamble \ort{ū́ta} / \ort{uúta}//
\gla u-tə//
\glb \gl{1}\gl{s_a_}-go//
\glft \qu{I go.}//
\endgl
\a<bak-35>
\begingl
\glpreamble \ort{töre-w-akine}//
\gla tərə w-a-kɨne//
\glb there \gl{1}\gl{s_a_}-be-\gl{pst}.\gl{cont}//
\glft \qu{I was there.}//
\endgl
\end{multicols}
\xe

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs with \gl{1}\gl{s_a_} \obj{w-} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
While there are reflexes of \rc{a(p)} \qu{to be} in \ikpeng, only \rc{eti} forms are attested inflected for first person \parencite[401]{gildea2018reconstructing}.

@@

%\ex<ikp-182>\ikpeng \parencite[][320]{chagas2019ikpeng}\\
%\begingl
%\gla ∅-iptoŋ-lan//
%\glb \gl{3}\gl{s_a_}-sit.down-\gl{pst}//
%\glft \qu{She sat down.}//
%\endgl
%\xe
%
While clearly not a regular sound change, \PXin \rc{w} is attested as sometimes being lost in \ikpeng, as evidenced by the correspondences in \exref{ikpwloss}.
Thus, we suggest that the \ikpeng \gl{1}\gl{s_a_} marker \obj{ɨ-}/Ø is cognate with \arara \gl{1}\gl{s_a_} \obj{w(ɨ)-}, and was subject to phonological erosion of \rc{w}, a phenomenon which will also be seen in other languages discussed in following sections.

@@

\begin{table}
	\centering
	\caption{Verbs preserving \gl{1}\gl{s_a_} in \PPek}
	\label{tab:pekverbs}
	\begin{tabular}{@{}lllll@{}}
	\mytoprule
& \PPek & \arara & \ikpeng & \bakairi \\
\mytoprule
\qu{say} & \rc{wɨ-ge-} & \obj{wɨ-ge-} & \obj{ɨ-ge-} & \obj{u-ge-} \\
\qu{bathe} & \rc{w-ipɨ-} & \obj{w-ibɨ-} & \obj{Ø-ip-} & \obj{w-i-} \\
\qu{be-1} & \rc{w-ap-} & \obj{w-ap-} & -- & \obj{w-a-} \\
\qu{be-2} & \rc{w-et͡ʃi-} & \obj{w-it͡ʃi-} & \obj{Ø-et͡ʃi-} & \obj{w-i-} \\
\qu{come} & \rc{w-epɨ}, \rc{k-əd-epɨ-} & \obj{w-ebɨ-} & \obj{k-arep-} & \obj{k-əewɨ-} \\
\qu{go down} & \rc{w-[ɨ/i]ptə-} & \obj{w-iptoŋ-} & ?\obj{-iptoŋ-} & \obj{k-ɨtəgɨ-} \\
\qu{go} & \rc{w-ɨtən-} & \obj{w-ɨdo-} & \obj{k-aran-} & \obj{u-tə-} \\
	\mybottomrule
	\end{tabular}
\end{table}

\Cref{tab:pekverbs} gives an overview of the first person forms of the seven verbs under discussion, along with our \PPek reconstruction.
The form for \rc{ge} \qu{to say} is straightforward to reconstruct; the \bakairi change of \rc{wɨ-} to \obj{u-} is not unexpected given correspondences like \bakairi \obj{udo} \parencite{meira2005southern} from \PC \rc{wɨtoto} \qu{person} \parencite[4]{gildea2007greenberg}.
\qu{to bathe} is straightforwardly reconstructible as \rc{ipɨ}.
For \qu{to be}, \ikpeng \obj{e} is very likely the original vowel, given the \PC form \rc{eti} \pcref{sec:be}.

@@

\ex<ara-123>\arara \parencite[][150]{alves2017arara}\\
\begingl
\gla m-odebɨ-nɨ//
\glb \gl{2}\gl{s_a_}-come-\gl{rec}//
\glft \qu{You came.}//
\endgl
\xe
%
We can imagine two scenarios leading to this situation:
One possibility is that there was some variation in \PPar as to whether or not the first person form of \qu{to go} was based on \rc{ətepɨ} or \rc{epɨ}.
Alternatively, the \PPek paradigm had a similar person-based distribution of \rc{əd-} as \arara, and \bakairi and \ikpeng both independently regularized the paradigm to show \rc{əd-epɨ} with all persons.
However, in both scenarios there is a correlation of \rc{w-} and \rc{ebɨ}, and \rc{k-} and \rc{əd-ebɨ}.

The forms for \qu{to go down} are also not fully cognate; \textcite{meira2005southern} make no mention of a regular correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}.
Rather, it seems that some roots gained a final \obj{ŋ} in \PXin, for so far unclear reasons: \begin{inlinelist}
\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} \parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} \parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} \parencites[8]{gildea2007greenberg}[57]{alves2017arara}
\end{inlinelist}.
Also, based on the fact \bakairi has generally lost much segmental material, we suggest that \bakairi \obj{gɨ} is yet another addition to a root \rc{ɨptə} or \rc{iptə}, rather than a conservative form.

The forms for \qu{to go} are all cognate, but while \arara and \bakairi point to \rc{ɨtə}, the \ikpeng form seems rather divergent.
However, it is in fact compatible with a reconstruction \rc{ɨtən}, when one considers that \ikpeng \obj{a} is an attested outcome of \rc{ə} (\PXin \rc{o}/\obj{e}), for example \obj{akari} \qu{dog} in \exref{ikpwloss} above, \obj{anma} \qu{path} \parencite[24]{ikpengpacheco2001} from \PC \rc{ətema} \parencite[12]{gildea2007greenberg}, or \obj{jaj} \qu{tree} from \PC \rc{jəje}.
This attested change of \rc{ə} to \obj{a} need only be preceded by a assimilatory lowering of initial \rc{ɨ} to \rc{ə}, to yield the form \obj{aran} from \rc{ɨtən}.
Other \ikpeng reflexes of \qu{to go} offer evidence for the suggested intermediate stage \rc{ətən}: \obj{ero-lɨ} \qu{s/he went} \parencite[25]{ikpengpacheco2001}.

Summing up, an innovative \gl{1}\gl{s_a_} marker \rc{k-} is reconstructible to \PPek.
Seven verbs can be reconstructed as having resisted this innovation, and preserving \gl{1}\gl{s_a_} \rc{w(ɨ)-} in \PPek.
For \qu{to come}, two distinct forms \rc{w-epɨ-} and \rc{k-əd-epɨ-} can be reconstructed, although the latter may be two parallel developments in \bakairi and \ikpeng.
In later, individual developments, \bakairi introduced \obj{k-} to \qu{to go down}, and \ikpeng to \qu{to go}.

@@

\ex<wais> Waiwaian \rc{eɸurka} \qu{to fall} (\gl{s_a_}) and \rc{wɨnɨkɨ} \qu{to sleep} (\gl{s_p_})\\
\begin{tabular}[t]{@{}lllllll@{}}
& \multicolumn{2}{l}{\PWai} & \multicolumn{2}{l}{\hixka}  & \multicolumn{2}{l}{\waiwai} \\
 & \gl{s_a_} & \gl{s_p_} & \gl{s_a_} & \gl{s_p_} & \multicolumn{2}{l}{\gl{s}} \\
\gl{1} & \rc{\emp{k-}eɸurka-} & \rc{kɨ-wɨnɨkɨ-} & \obj{\emp{k-}ehurka-} & \obj{kɨ-nɨkɨ-} & \obj{\emp{k-}eɸɨrka-} & \obj{kɨ-wɨnk-} \\
\gl{2} & \rc{m-eɸurka-} & \rc{o-wɨnɨkɨ-} & \obj{m-ehurka-} & \obj{o-wnɨkɨ-} & \obj{m-eɸɨrka-} & \obj{mɨ-wɨnk-} \\
\gl{1+2} & \rc{t-eɸurka-} & \rc{tɨt-wɨnɨkɨ-} & \obj{t-ehurka-} & \obj{tɨ-nɨkɨ-} & \obj{t͡ʃ-eɸɨrka-} & \obj{tɨt-wɨnk-} \\
\gl{3} & \rc{ɲ-eɸurka-} & \rc{nɨ-wɨnɨkɨ-} & \obj{ɲ-ehurka-} & \obj{nɨ-nɨkɨ-} & \obj{ɲ-eɸɨrka-} & \obj{nɨ-wɨnk-} \\
\end{tabular}\\
\parencites[510]{howard2001wrought}[189--191]{hixkaryanaderby1985}[209--211]{hawkins1953waiwai}[50]{waiwaihawkins1998}
\xe
%
\waiwai \obj{ka} \qu{to say} does not take \obj{kɨ-}, but rather conservative \obj{wɨ-} \exref{hixɨ.wai-108}.
Its \hixka counterpart also takes a reflex of \gl{1}\gl{s_a_} \rc{w-}, which has the form \obj{ɨ-} \exref{hixɨ.hix-118}.
A formally identical prefix is found on transitive verbs in \gl{1}>\gl{3} scenarios \exref{hixɨ.hix-17}, where it also corresponds to \waiwai \obj{w(ɨ)-} \exref{hixɨ.wai-159}.

\pex<hixɨ>
\a<wai-108> \waiwai \parencite[][71]{waiwaihawkins1998}
\begingl
\glpreamble wɨɨkekɲe//
\gla wɨ-ka-jakɲe//
\glb \gl{1}-say-\gl{pst}//
\glft \qu{I said.}//
\endgl
\a<hix-118> \hixka \parencite[][124]{hixkaryanaderby1985}
\begingl
\glpreamble roxehra nay hamɨ Kaywerye ɨkekonɨ//
\gla ro-ʃe-hɨra n-a-je hamɨ kajwerʲe ɨ-ka-jakonɨ//
\glb \gl{1}-\gl{des}-\gl{neg} \gl{3}-be-\gl{npst}.\gl{uncert} \gl{evid} K. \gl{1}\gl{s_a_}-say-\gl{rem}.\gl{cont}//
\glft \qu{I said (to myself), “Kaywerye evidently doesn't like me”.}//
\endgl
\a<hix-17> \hixka \parencite[][191]{hixkaryanaderby1985}
\begingl
\gla ɨ-koroka-no//
\glb \gl{1}>\gl{3}-wash-\gl{imm}//
\glft \qu{I washed him.}//
\endgl
\a<wai-159> \waiwai \parencite[][192]{waiwaihawkins1998}
\begingl
\glpreamble wîyesî//
\gla wɨ-jo-jasɨ//
\glb \gl{1}>\gl{3}-boil-\gl{npst}//
\glft \qu{I will boil it.}//
\endgl
\xe
%
Thus, similar to the the case of \ikpeng, \hixka has lost the \rc{w} of the \envr{}{C} allomorph \rc{wɨ-}.
Notably, \textcite{hixkaryanaderby1985} does not see this \obj{ɨ-} as an irregular \gl{1}\gl{s_a_}, but as the regular \gl{1}>\gl{3} marker, because he considers \hixka \obj{ka} \qu{to say} to be transitive, an issue which will be discussed in \cref{sec:say}.

There are three more verbs which did not take innovative \rc{k-} in \PWai.
Two roots are part of the copula \rc{eʃi}/\obj{a(h)}, straightforwardly reconstructed in \exref{parpar.parcop}.
The other is \qu{to go} \exref{parpar.pargo}, somewhat of a special case.

@@

Summing up, we reconstruct the three verbs \rc{et͡ʃi}/\obj{a(h)} \qu{to be}, \rc{ka(s)} \qu{to say}, and \rc{to} \qu{to go} as preserving the old \gl{1}\gl{s_a_} marker \rc{w-} in \PWai, while the rest took on innovative \rc{k-}.

\subsection{\PTir \rc{t-}}
\label{sec:taranoan}
The Taranoan languages were already grouped together by \textcite{girard1971proto}, consisting of closely related \trio and \akuriyo (subsumed here under the moniker Tiriyoan), and the more distant member \carijo.
\textcite{meira1998proto} provides an extensive phonological, morphological, and lexical reconstruction of \PTar.
The \setone paradigms of \trio and \akuriyo contain an interesting puzzle: \PC \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} \rc{w-} seem to have switched places.
%This is illustrated by comparing reconstructible \PTir forms with more conservative \wayana \exref{sat} and \kalina \exref{13w}, which can be taken as representative for \PC.
%
%\ex<sat>
%\begin{tabular}[t]{@{}llllll@{}}
%\gl{1}\gl{s_a_} & \PC & \wayana & \PTir & \trio & \akuriyo\\
%\qu{yawn} & \rc{w-e-mɨta-po-} & \obj{w-emtapɨ-} & \rc{t͡ʃ-entapo-} & \obj{s-entapo-} & \obj{t͡ʃ-entapo-}\\
%\qu{bathe} & \rc{w-e-pɨ-} & \obj{w-epɨ-} & \rc{t͡ʃ-epɨ-} & \obj{s-epɨ-} & \obj{t͡ʃ-epɨ-}\\
%\qu{talk} & \rc{w-ət-uru} & \obj{w-əturu-} & \rc{t-əturu-} & \obj{t-əturu-} & \obj{k-əturu-}\footnotemark\\
%\qu{learn} & \rc{w-ət-eme-pa} & \obj{w-əhepa-} & \rc{t-əempa-}\footnotemark & \obj{t-əenpa-} & \obj{k-əempa-}\\
%\end{tabular}\\
%\parencites[28]{camargo2010wayana}[292, 664]{triomeira1999}[77, 86, 82]{gildea1994akuriyo}[50, 208, 213]{wayanatavares2005}
%\xe
%\addtocounter{footnote}{-1}
%\footnotetext{\PTir \rc{t-} was subsequently replaced by \obj{k-} in \akuriyo \pcref{sec:akuriyo}.}
%\addtocounter{footnote}{1}
%\footnotetext{
%The \trio form is not due to dissimilation from \rc{mp} to \obj{np}, but rather a phonemic transcription rendering the syllable-final archiphoneme \pnm{N} as \obj{n} \parencite[31]{triomeira1999}.
%Further, \PTar intervocalic \rc{t͡ʃ} was lost entirely in \trio (except after \rc{i}) and is reflected as \obj{ʔ} in \akuriyo \parencite[31--32]{meira1998proto}.
%Given the original shape of the detransitivizers \rc{ət(e)-} and \rc{e-}, \trio \obj{ə-enpa} \qu{to learn} contains a reflex of the former with loss of \rc{t}, so \akuriyo \obj{əempa} would be expected to have a glottal stop: ?\obj{əʔempa}.
%Its absence could either be due to \trio influence or not being transcribed by \textcite{gildea1994akuriyo}.

>>>>

\ex<bak-33>\bakairi \parencite[][4]{meira2003bakairi}\\
\begingl
\gla w-i-də//
\glb \gl{1}\gl{s_a_}-bathe-\gl{imm}//
\glft \qu{I bathed}//
\endgl
\xe
%
While \textcite[4]{meira2003bakairi} does list \obj{ge} \qu{to say}, \obj{tə} \qu{to go}, and \obj{əe(wɨ)} \qu{to come} as examples of \gl{s_a_} verbs, he does not indicate whether they belong to the class of \gl{s_a_}-1 verbs, with first person \obj{k-}, or the \gl{s_a_}-2 verbs, with \obj{w-}.\footnote{It should be noted at this point that \textcite{meira2003bakairi} indicates that the same verbs which take first person \obj{w-} in \bakairi also take a \gl{1+2} marker \obj{k-}. However, this marker is only illustrated for \qu{to bathe}, both by \textcite{meira2003bakairi} and \textcite{von1892bakairi}. Given the lack of data for other verbs, we will not further discuss this potential additional pattern. If the characterization by \citeauthor{meira2003bakairi} is accurate, then the pattern is fully parallel to the distribution of the first person prefixes.}
Luckily, while \textcite{von1892bakairi} did not accurately record all phonemic distinctions in \bakairi \parencite{meira2005bakairi}, he does provide inflected forms of cognates to the \arara verbs in \exref{arairr}.
We present them according to the phonemic analysis of \bakairi by \textcites{meira2003bakairi}{meira2005bakairi} in \exref{bakverbs}.

\pex[everyglpreamble=]<bakverbs> \bakairi \parencite[][131, 397, 76, 137, 374]{von1892bakairi}
\begin{multicols}{2}
\a<bak-3>
\begingl
\glpreamble \ort{u-ɣépa} //
\gla u-ge-pa//
\glb \gl{1}\gl{s_a_}-say-\gl{neg}//
\glft \qu{I don't say.}//
\endgl
\a<bak-5>
\begingl
\glpreamble \ort{wi-táki} / \ort{wi-tági} //
\gla w-i-taki//
\glb \gl{1}\gl{s_a_}-be-\gl{int}//
\glft \qu{I was.}//
\endgl
\a<bak-4>
\begingl
\glpreamble \ort{kχaewí-le} //
\gla k-əewɨ-lɨ//
\glb \gl{1}\gl{s_a_}-come-\gl{imm}//
\glft \qu{I came.}//
\endgl
\a<bak-6>
\begingl
\glpreamble \ort{kχ-itaké-he} //
\gla k-ɨtəgɨ-se//
\glb \gl{1}\gl{s_a_}-go.down-\gl{npst}?//
\glft \qu{I go down.}//
\endgl
\a<bak-2>
\begingl
\glpreamble \ort{ū́ta} / \ort{uúta}//
\gla u-tə//
\glb \gl{1}\gl{s_a_}-go//
\glft \qu{I go.}//
\endgl
\end{multicols}
\xe

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs in \exref{arairr} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down} \exref{ikp-182}, but no first person forms are attested \perscommpar{Angela Chagas}.

@@

\ex<ikp-182>\ikpeng \parencite[][320]{chagas2019ikpeng}\\
\begingl
\gla ∅-iptoŋ-lan//
\glb \gl{3}\gl{s_a_}-sit.down-\gl{pst}//
\glft \qu{She sat down.}//
\endgl
\xe
%
While clearly not a regular sound change, \PXin \rc{w} is attested as sometimes being lost in \ikpeng, as evidenced by the correspondences in \exref{ikpwloss}.
Thus, we suggest that the \ikpeng \gl{1}\gl{s_a_} marker \obj{ɨ-}/Ø is cognate with \arara \gl{1}\gl{s_a_} \obj{w(ɨ)-}, and was subject to phonological erosion of \rc{w}, a phenomenon which can also be found in other languages discussed in following sections.

@@

\begin{table}
	\centering
	\caption{Verbs preserving \gl{1}\gl{s_a_} in \PPek}
	\label{tab:pekverbs}
	\begin{tabular}{@{}lllll@{}}
	\mytoprule
& \PPek & \arara & \ikpeng & \bakairi \\
\mytoprule
\qu{say} & \rc{wɨ-ge-} & \obj{wɨ-ge-} & \obj{ɨ-ge-} & \obj{u-ge-} \\
\qu{bathe} & \rc{w-ipɨ-} & \obj{w-ibɨ-} & \obj{Ø-ip-} & \obj{w-i-} \\
\qu{be} & \rc{w-et͡ʃi-} & \obj{w-it͡ʃi-} & \obj{Ø-et͡ʃi-} & \obj{w-i-} \\
\qu{come} & \rc{w-epɨ}, \rc{k-əd-epɨ-} & \obj{w-ebɨ-} & \obj{k-arep-} & \obj{k-əewɨ-} \\
\qu{go down} & \rc{w-[ɨ/i]ptə-} & \obj{w-iptoŋ-} & ?\obj{-iptoŋ-} & \obj{k-ɨtəgɨ-} \\
\qu{go} & \rc{w-ɨtən-} & \obj{w-ɨdo-} & \obj{k-aran-} & \obj{u-tə-} \\
	\mybottomrule
	\end{tabular}
\end{table}

\Cref{tab:pekverbs} gives an overview of the first person forms of the six verbs under discussion, along with our \PPek reconstruction.
The form for \rc{ge} \qu{to say} is straightforward to reconstruct; the \bakairi change of \rc{wɨ-} to \obj{u-} is not unexpected given correspondences like \bakairi \obj{udo} \parencite{meira2005southern} from \PC \rc{wɨtoto} \qu{person} \parencite[4]{gildea2007greenberg}.
\qu{to bathe} is straightforwardly reconstructible as \rc{ipɨ}.
For \qu{to be}, \ikpeng \obj{e} is very likely the original vowel, given the \PC form \rc{eti} \pcref{sec:be}.

@@

\ex<ara-123>\arara \parencite[][150]{alves2017arara}\\
\begingl
\gla m-odebɨ-nɨ//
\glb \gl{2}\gl{s_a_}-come-\gl{rec}//
\glft \qu{You came.}//
\endgl
\xe
%
Thus, it seems that there was some variation in \PPar as to whether or not the first person form of \qu{to go} was based on \rc{ətepɨ} or \rc{epɨ}.
Given that \bakairi and \ikpeng agree about both the first person prefix and the reflex of \rc{ət-}, we reconstruct two distinct first person forms, \rc{w-epɨ-} and \rc{k-ət-epɨ-}.\footnote{It may also be the case that the \PPek paradigm had a similar person-based distribution of \rc{əd-} as \arara, and \bakairi and \ikpeng both independently regularized the paradigm to show \rc{əd-epɨ} with all persons. If that were the case, the connection between \rc{k-} and \rc{əd-epɨ} would still exist.}

The forms for \qu{to go down} are again not fully cognate; \textcite{meira2005southern} make no mention of a regular correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}.
Rather, it seems that some roots gained a final \obj{ŋ} in \PXin, for so far unclear reasons: \begin{inlinelist}
\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} \parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} \parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} \parencites[8]{gildea2007greenberg}[57]{alves2017arara}
\end{inlinelist}.
Also, based on the fact \bakairi has generally lost much segmental material, we suggest that \bakairi \obj{gɨ} is yet another addition to a root \rc{ɨptə} or \rc{iptə}, rather than a conservative form.

The forms for \qu{to go} are all cognate, but while \arara and \bakairi point to \rc{ɨtə}, the \ikpeng form is rather divergent.
However, it is in fact compatible with a reconstruction \rc{ɨtən}, when one considers that \ikpeng \obj{a} is an attested outcome of \rc{ə} (\PXin \rc{o}/\obj{e}), for example \obj{akari} \qu{dog} in \exref{ikpwloss} above, \obj{anma} \qu{path} \parencite[24]{ikpengpacheco2001} from \PC \rc{ətema} \parencite[12]{gildea2007greenberg}, or \obj{jaj} \qu{tree} from \PC \rc{jəje}.
This attested change of \rc{ə} to \obj{a} need only be preceded by a assimilatory lowering of initial \rc{ɨ} to \rc{ə}, to yield the form \obj{aran} from \rc{ɨtən}.
Other \ikpeng reflexes of \qu{to go} offer evidence for the suggested intermediate stage \rc{ətən}: \obj{ero-lɨ} \qu{s/he went} \parencite[25]{ikpengpacheco2001}.

Summing up, an innovative \gl{1}\gl{s_a_} marker \rc{k-} is reconstructible to \PPek.
Seven verbs can be reconstructed as having resisted this innovation, and preserving \gl{1}\gl{s_a_} \rc{w(ɨ)-} in \PPek.
For \qu{to come}, two distinct forms \rc{w-epɨ-} and \rc{k-əd-epɨ-} must be reconstructed.
In later, individual developments, \bakairi introduced \obj{k-} to \qu{to go down}, and \ikpeng to \qu{to go}.

@@

\ex<wais> Waiwaian \rc{eɸurka} \qu{to fall} (\gl{s_a_}) and \rc{wɨnɨkɨ} \qu{to sleep} (\gl{s_p_})\\
\begin{tabular}[t]{@{}lllllll@{}}
& \multicolumn{2}{l}{\PWai} & \multicolumn{2}{l}{\hixka}  & \multicolumn{2}{l}{\waiwai} \\
 & \gl{s_a_} & \gl{s_p_} & \gl{s_a_} & \gl{s_p_} & \multicolumn{2}{l}{\gl{s}} \\
\gl{1} & \rc{\emp{k-}eɸurka-} & \rc{kɨ-wɨnɨkɨ-} & \obj{\emp{k-}ehurka-} & \obj{kɨ-nɨkɨ-} & \obj{\emp{k-}eɸɨrka-} & \obj{kɨ-wɨnk-} \\
\gl{2} & \rc{m-eɸurka-} & \rc{o-wɨnɨkɨ-} & \obj{m-ehurka-} & \obj{o-wnɨkɨ-} & \obj{m-eɸɨrka-} & \obj{mɨ-wɨnk-} \\
\gl{1+2} & \rc{t-eɸurka-} & \rc{tɨt-wɨnɨkɨ-} & \obj{t-ehurka-} & \obj{tɨ-nɨkɨ-} & \obj{t͡ʃ-eɸɨrka-} & \obj{tɨt-wɨnk-} \\
\gl{3} & \rc{ɲ-eɸurka-} & \rc{nɨ-wɨnɨkɨ-} & \obj{ɲ-ehurka-} & \obj{nɨ-nɨkɨ-} & \obj{ɲ-eɸɨrka-} & \obj{nɨ-wɨnk-} \\
\end{tabular}\\
\parencites[510]{howard2001wrought}[189--191]{hixkaryanaderby1985}[209--211]{hawkins1953waiwai}[50]{waiwaihawkins1998}
\xe
%
The \waiwai verb \obj{ka} \qu{to say} was already shown have resisted this extension in \cref{sec:extensions_intro}, showing \obj{wɨ-}, a reflex of the \PC \gl{1}\gl{s_a_} marker \rc{w-}.
Its \hixka counterpart also takes a reflex of \gl{1}\gl{s_a_} \rc{w-}, which has the form \obj{ɨ-} \exref{hixɨ.hix-118}.
A formally identical prefix is found on transitive verbs in \gl{1}>\gl{3} scenarios \exref{hixɨ.hix-17}, where it also corresponds to \waiwai \obj{w(ɨ)-} \exref{hixɨ.wai-159}.

\pex<hixɨ>
\a<hix-118> \hixka \parencite[][124]{hixkaryanaderby1985}\\
\begingl
\glpreamble roxehra nay hamɨ Kaywerye ɨkekonɨ//
\gla ro-ʃe-hɨra n-a-je hamɨ kajwerʲe ɨ-ka-jakonɨ//
\glb \gl{1}-\gl{des}-\gl{neg} \gl{3}-be-\gl{npst}.\gl{uncert} \gl{evid} K. \gl{1}\gl{s_a_}-say-\gl{rem}.\gl{cont}//
\glft \qu{I said (to myself), “Kaywerye evidently doesn't like me”.}//
\endgl
\a<hix-17> \hixka \parencite[][191]{hixkaryanaderby1985}\\
\begingl
\gla ɨ-koroka-no//
\glb \gl{1}>\gl{3}-wash-\gl{imm}//
\glft \qu{I washed him.}//
\endgl
\a<wai-159> \waiwai \parencite[][192]{waiwaihawkins1998}\\
\begingl
\glpreamble wîyesî//
\gla wɨ-jo-jasɨ//
\glb \gl{1}>\gl{3}-boil-\gl{npst}//
\glft \qu{I will boil it.}//
\endgl
\xe
%
Thus, similar to the the case of \ikpeng, \hixka has lost the \rc{w} of the \envr{}{C} allomorph \rc{wɨ-}.
Notably, \textcite{hixkaryanaderby1985} does not see this \obj{ɨ-} as an irregular \gl{1}\gl{s_a_}, but as the regular \gl{1}>\gl{3} marker, because he considers \hixka \obj{ka} \qu{to say} to be transitive, an issue which will be discussed in \cref{sec:say}.

There are two more verbs which did not take innovative \rc{k-} in \PWai.
One of them is the copula \rc{eʃi}/\obj{a(h)}, reconstructed in \exref{parpar.parcop}.
The other is \qu{to go} \exref{parpar.pargo}, somewhat of a special case.

@@

Summing up, I reconstruct the three verbs \rc{et͡ʃi}/\obj{a(h)} \qu{to be}, \rc{ka(s)} \qu{to say}, and \rc{to} \qu{to go} as preserving the old \gl{1}\gl{s_a_} marker \rc{w-} in \PWai, while the rest took on innovative \rc{k-}.

\subsection{\PTir \rc{t-}}
\label{sec:taranoan}
The Taranoan languages were already grouped together by \textcite{girard1971proto}, consisting of closely related \trio and \akuriyo (subsumed here under the moniker Tiriyoan), and the more distant member \carijo.
\textcite{meira1998proto} provides an extensive phonological, morphological, and lexical reconstruction of \PTar.
The \setone paradigms of \trio and \akuriyo contain an interesting puzzle: \PC \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} \rc{w-} seem to have switched places.
This is illustrated by comparing reconstructible \PTir forms with more conservative \wayana \exref{sat} and \kalina \exref{13w}, which can be taken as representative for \PC.

\ex<sat>
\begin{tabular}[t]{@{}llllll@{}}
\gl{1}\gl{s_a_} & \PC & \wayana & \PTir & \trio & \akuriyo\\
\qu{yawn} & \rc{w-e-mɨta-po-} & \obj{w-emtapɨ-} & \rc{t͡ʃ-entapo-} & \obj{s-entapo-} & \obj{t͡ʃ-entapo-}\\
\qu{bathe} & \rc{w-e-pɨ-} & \obj{w-epɨ-} & \rc{t͡ʃ-epɨ-} & \obj{s-epɨ-} & \obj{t͡ʃ-epɨ-}\\
\qu{talk} & \rc{w-ət-uru} & \obj{w-əturu-} & \rc{t-əturu-} & \obj{t-əturu-} & \obj{k-əturu-}\footnotemark\\
\qu{learn} & \rc{w-ət-eme-pa} & \obj{w-əhepa-} & \rc{t-əempa-}\footnotemark & \obj{t-əenpa-} & \obj{k-əempa-}\\
\end{tabular}\\
\parencites[28]{camargo2010wayana}[292, 664]{triomeira1999}[77, 86, 82]{gildea1994akuriyo}[50, 208, 213]{wayanatavares2005}
\xe
\addtocounter{footnote}{-1}
\footnotetext{\PTir \rc{t-} was subsequently replaced by \obj{k-} in \akuriyo \pcref{sec:akuriyo}.}
\addtocounter{footnote}{1}
\footnotetext{
The \trio form is not due to dissimilation from \rc{mp} to \obj{np}, but rather a phonemic transcription rendering the syllable-final archiphoneme \pnm{N} as \obj{n} \parencite[31]{triomeira1999}.
Further, \PTar intervocalic \rc{t͡ʃ} was lost entirely in \trio (except after \rc{i}) and is reflected as \obj{ʔ} in \akuriyo \parencite[31--32]{meira1998proto}.
Given the original shape of the detransitivizers \rc{ət(e)-} and \rc{e-}, \trio \obj{ə-enpa} \qu{to learn} contains a reflex of the former with loss of \rc{t}, so \akuriyo \obj{əempa} would be expected to have a glottal stop: ?\obj{əʔempa}.
Its absence could either be due to \trio influence or not being transcribed by \textcite{gildea1994akuriyo}.
Since there is no evidence for a glottal stop in \PTir (\rc{əʔempa}), I do not reconstruct it.}
